---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
---


```{r, echo=FALSE}
options(tidyverse.quiet = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))
#suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(glue))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x, digits=2){

  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
      paste(format(x,big.mark=',', digits=digits, nsmall=digits, scientific=FALSE))
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)

package_version <- function(package){

  paste(unlist(packageVersion(package)), collapse='.')

}


oxford_comma <- function(x, digits=2) {

  x <- map_chr(x, inline_hook, digits=digits)

  if(length(x) < 2){
    x
  } else if(length(x) == 2){
    paste(x, collapse = " and ")
  } else {
    paste(paste(x[-length(x)], collapse=", "), x[length(x)], sep=", and ")
  }
}

capitalize <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}


datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
              "rainforest", "rice", "seagrass", "sediment", "soil", "stream")

pretty_datasets <- tibble(
  plain = datasets,
  pretty = capitalize(plain)
)

```

# Waste not, want not: Revisiting the analysis that called rarefaction into question

\vspace{20mm}

**Running title:** Review of "Waste not, want not"

\vspace{20mm}

Patrick D. Schloss${^\dagger}$

\vspace{40mm}

${\dagger}$ To whom corresponsdence should be addressed:


\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology & Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

**Research article**

\newpage

```{r datasets}

```

## Abstract

\newpage

## Introduction

## Results

## Discussion

## Acknowledgements

\newpage

## References

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>
\bibliography{ref}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

## Figures
